---
output: html_document
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo=FALSE,warning = FALSE)
```
#Calculate background 21Ne from U and Th concentrations
```{r,include=FALSE}

#Remove all lists
rm(list=ls())

#Load libraries
library(tidyverse)
library(stringr)
library(DT)
```

#Read original data csv from Dan Morgan

```{r}
my_data <- tbl_df(read.csv("~/Google Drive/U-Th_in_Ne-21/GeochemistryDataAntML2015.csv" , stringsAsFactors = FALSE)) %>% 
  rename(sample_id = ANALYTE, wtkg = WtKg)
my_data
```
#Clean up the data via tidy data
```{r}
#set row names to sample ID's

#get meta data from csv cells
header_names <- my_data$sample_id[1:3]
meta = my_data %>% 
  filter(sample_id %in% header_names)

# take my_data, call it true_data
true_data <- my_data %>% 
  filter(!(sample_id %in% header_names)) %>% 
  as_data_frame() %>% 
  gather(element, ppm, -sample_id,-wtkg) %>% 
  mutate(numeric_ppm = as.numeric(ppm),
         wtkg = as.numeric(wtkg), 
         below_limit_ppm = str_detect(ppm,"<")) %>%
  arrange((sample_id)) 


#Do any operation to all entries titled sample_id
#this is an example
true_data %>% 
  group_by(sample_id) %>% 
  summarise(domath=numeric_ppm[element=="U"]*numeric_ppm[element=="Th"],
            num_elements = sum(is.na(numeric_ppm) == F))
true_data
```
#Add Constants:
```{r}
#Analytical Resuts
sampleAge = 100e6 #Ma

#Standards and Constants
U238_abundance = 0.992742
U235_abundance = 0.007204
Th232_abundance = 1
abundance = c(U238_abundance,U235_abundance,Th232_abundance)
#names(abundance)<- elements

#Isotope constants
U238_mass= 238.050788247
U235_mass = 235.043929918
Th232_mass = 232.038055325
mass = c(U238_mass,U235_mass,Th232_mass)
#names(mass)<- elements

#Number of alpha decays for the radioactive reaction series. (online)
U238_num_alphas = 8
U235_num_alphas = 7
Th232_num_alphas = 6
num_alpha_decays_in_chain = c(U238_num_alphas,U235_num_alphas,Th232_num_alphas)
#names(num_alphas)<-elements

avogadro= 6.022E23

#Nucleogenic Neon Production Rates
U238_21Ne_Prate = 4.04E-8 #via Table 6, Coxx et al 2015
U235_21Ne_Prate = 5.62E-8 #via Table 6, Coxx et al 2015
Th232_21Ne_Prate = 6.08E-8 #via Table 6, Coxx et al 2015
ne21_nucleogenic_P_rate = c(U238_21Ne_Prate,U235_21Ne_Prate,Th232_21Ne_Prate)
#names(Ne21_Nucl_Prate)<- elements


#Alpha Decay Constants "lambda"= ln(2)/T_halflife (confirmed 09/21/2016)
U238_alpha = log(2)/4.468E9
U235_alpha = log(2)/7.038E8
Th232_alpha = log(2)/1.405E10
lambda = c(U238_alpha,U235_alpha,Th232_alpha)
#names(lambda)<- elements

# Create a data frame of the constants
elem_info=data_frame(element=c("U","U","Th"),isotope=c("U238","U235","Th232"),abundance,
mass,num_alpha_decays_in_chain,ne21_nucleogenic_P_rate,lambda)
```
#Split elements into U and Th isotopes
```{r}
#-----write a function to split elements into isotopes U235 and U238-------------
#so we can create a big ole data frame in which to do parrallel math across rows

elements <- unique(true_data$element)

atom_splitter=function(i, dat, elem_info){
  
  tdat <- dat %>% 
    filter(element == i) %>% 
    mutate(isotope = NA,
           iso_ppm = numeric_ppm,
           isoprop = 1)
  
  isos <- elem_info %>% 
    filter(element == i)
  
  if (dim(isos)[1] > 0){
    
    props <- rep(isos$abundance,length(unique(tdat$sample_id)))
    
    isotopes <-  rep(isos$isotope,length(unique(tdat$sample_id)))
      
    tdat <- tdat[rep(1:dim(tdat)[1], length(isos$isotope)),] %>% 
      arrange(sample_id) %>% 
      mutate(isoprop = props, 
             iso_ppm = numeric_ppm * isoprop,
             isotope = isotopes)
    
  }
  
  return(tdat)
}

extended_elements <- lapply(elements, atom_splitter, dat = true_data, elem_info = elem_info) %>% 
  bind_rows()

extended_elements
```
#Merge data and constants into single dataframe
```{r}
#So first get everything in a single
#merge the two df's based on common obervation known as isotope
extended_elements=full_join(extended_elements,elem_info,by="isotope")
#format a clean dataframe
extended_elements=tbl_df(extended_elements)

#ok! now we have a big clean data set in which to do cross math with!
#BUT... need to add other constants like avogadro and sample age
extended_elements=mutate(extended_elements, avogadro,sampleAge)

datatable(extended_elements)
```
# Analyze data
```{r}
# Ok so the data is all there finally, now do some math across columns
#filter entries that don't have isotope abundances since you can't so math on "NA's"

extended_elements=extended_elements %>% 
  filter(is.na(abundance)==FALSE ) %>% 
  mutate(parent_atoms = (numeric_ppm/1E6)*wtkg*1000*abundance*(1/mass)*avogadro) %>% 
  as_data_frame()

```
#Calculate # of alpha decays given age of sample
```{r}
extended_elements=extended_elements %>% 
  mutate(alpha_decays = parent_atoms*(exp(lambda*sampleAge)-1))

#Plot decays/sample
extended_elements %>% 
  ggplot(aes(x=sample_id,y=alpha_decays, color=isotope))+
  geom_point()+
  geom_smooth()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ggtitle("# of Alpha decays based on age of sample")+
  ylab("# of alpha decays")
```
#Calculate alpha particle emitted over time given a U and Th concentration in sample
```{r}
extended_elements=extended_elements %>% 
  mutate(alpha_particles = num_alpha_decays_in_chain*alpha_decays)
#Plot alpha particles/decay
extended_elements %>% 
  ggplot(aes(x=alpha_particles,y=alpha_decays, color=isotope))+
  geom_point()+
  geom_smooth()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ggtitle("# of alpha particles vs alpha decays")+
  xlab("# of alpha particles")+
  ylab("# of alpha decays")
```
#Calculate # of 21Ne atoms produced from isotope decays due to age of sample
This is the number of background 21ne in a sample due to the concentration of U and Th
```{r}
extended_elements=extended_elements %>% 
  mutate(ne21_nucl_n_atoms = alpha_particles*ne21_nucleogenic_P_rate)

extended_elements %>% 
  ggplot(aes(x=ne21_nucl_n_atoms,y=numeric_ppm, color=isotope))+
  geom_point()+
  geom_smooth()+
  ggtitle("Nucleogenic 21Ne vs Th & U concentration")+
  xlab("Nucleogenic 21Ne (atoms)")+
  ylab("Concentration (ppm)")

#Try a histogram showing the porportion of each isotope per sample
extended_elements %>% 
  ggplot(aes(x=sample_id,y=ne21_nucl_n_atoms,fill = isotope))+
  geom_bar(stat = "identity")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ggtitle("Th & U fractions of Nucelogenic Neon by sample")+
  xlab("")+
  ylab("Nucleogenic 21Ne (atoms)")

```
#Plotting a dot for each sample and the amount of 21Ne in it, colored by the isotope
```{r}
extended_elements %>% 
  ggplot(aes(x=sample_id,y=ne21_nucl_n_atoms, color=isotope))+
  geom_point(aes(shape = isotope))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ggtitle("Th & U fractions of Nucelogenic Neon by sample")+
  xlab("")+
  ylab("Nucleogenic 21Ne (atoms)")
  #scale_y_log10()
```
#Calculate atoms per gram qtz
```{r}
extended_elements= extended_elements %>% 
  mutate(ne21_atoms_per_gram = ne21_nucl_n_atoms/wtkg)


```






